A North Sámi to South Sámi machine translation prototype

### 2014-02-09 paper/paper.tex
xxx

The bilingual dictionary: Some numbers 
\begin{itemize}
\item 3568 general word pairs + 61 proper nouns
\item added: 873 special domain word pairs 
\item added: 10 721 proper nouns
\end{itemize}

@@

The existing Norwegian-South Saami lexical resources are poor, with little terminology from the modern society. South Saami has been used more for giving information to the Saami people especially the last two years, and there are some texts especially about school administration and curriculum on Internet. By comparing Norwegian texts with the translations to North Saami and South Saami, we added 873 special domain word pairs to the bilingual dictionary.

Large syntactic differences also have consequences for the bilingual dictionary, because many North Saami verbs have to be translated into South Saami object+verb og adverb+verb. Examples are Norwegian `å trekke lodd' is translated into North Saami \textit{vuorbádit} and North Saami \textit{vuerpiem giesedh}.

@@

The pivot model causes some lexical problems. Even if the lexical inndeling often is similar in South Saami and North Saami, as in for the ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}. But in South Saami is no common verb for these three, and the lexical xxx is like in Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.

\subsubsection{Lexical selection}
Because we were tilpasse the MT system to a special domain, we have not made many rules for lexical selection. How many rules? But an MT-system for more domains, will certainly need more such rules.

>>>>

The bilingual dictionary: Some numbers

\begin{itemize}
\item 3568 general word pairs + 61 proper nouns
\item added: 873 special domain word pairs 
\item added: 10 721 proper nouns
\end{itemize}

@@

The existing Norwegian-South Saami lexical resources are poor, with little terminology from the modern society. South Saami has been used more for giving information to the Saami people especially during the last two years, and there are some texts especially about school administration and curriculum on Internet. By comparing Norwegian texts with the translations to North Saami and South Saami, we added 873 special domain word pairs to the bilingual dictionary.

Large syntactic differences also have consequences for the bilingual dictionary, because many North Saami verbs have to be translated into South Saami as $object+verb$ and $adverb+verb$ strings. Examples include Norwegian `å trekke lodd', («draw lots for») which is translated into North Saami as one verb \textit{vuorbádit}, but into South Saami as object+verb \textit{vuerpiem giesedh} (lit. «lot draw»).

@@

The pivot model causes some lexical problems. Even if the lexical conceptualisation often is similar in South Saami and North Saami, as for the trivial ‘to rain, to calcute’ pair, there are also counter examples, like the Norwegian verbs \textit{lese, telle, uttale} ‘read, count, utter’, which all three can be expressed with the North Saami verb \textit{lohkat}. But in South Saami there is no common verb for these concepts, and the lexical distribution is like the one for Norwegian: \textit{lohkedh, ryöknehtidh, jiehtedh}.

\subsubsection{Lexical selection}
Because we were adapting the MT system to a special domain, we have not made many rules for lexical selection. How many rules? But an MT-system for more domains, will certainly need more such rules.
